Channel Shaping Using Beyond Diagonal Reconfigurable Intelligent Surfaces

### 2024-05-15 main.tex
:triangular_flag_on_post: Bruno 240515

	Second, we propose an efficient \gls{bd}-\gls{ris} design framework that solves general optimization problems by geodesic\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} \gls{rcg}.
	% to numerically analyze the channel shaping capability of \gls{bd}-\gls{ris}, we exploit the Riemannian geometry of the Stiefel manifold and
	This method modified from \cite{Abrudan2008,Abrudan2009} exploits the Riemannian geometry of the Stiefel manifold and performs group-wise multiplicative rotational updates along the geodesics, which avoids retraction and facilitates step size selection.
	It thus converges faster than existing non-geodesic approaches \cite{Absil2009,Pan2022d}.
	This is the first work to tailor an efficient optimization framework for \gls{bd}-\gls{ris}.

@@

\begin{section}{\glsfmtshort{bd}-\glsfmtshort{ris} Design by Geodesic \glsfmtshort{rcg}}
	% In this section, we first provide an overview on signal processing techniques for group-connected \gls{bd}-\gls{ris} design problems, then propose a novel group-wise geodesic \gls{rcg} method that exploits the properties of unitary group and operates directly on the Stiefel manifold.
	% The proposed method is applicable to any smooth optimization problem on the Stiefel manifold and is particularly suitable for \gls{bd}-\gls{ris} design problems.
	% We will later show that the proposed method provides better objective value and faster convergence for smooth optimization problems of arbitrary group size.
	A geodesic is a curve representing the shortest path between two points in a Riemannian manifold, whose tangent vectors remain parallel when transporting along the curve.
	We notice that the feasible domain of scattering matrix of each \gls{bd}-\gls{ris} group is a $L$-dimensional Stiefel manifold $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$, which is non-convex and non-Euclidean.
	Therefore, relevant optimization problems are usually solved by relax-then-project methods \cite{Fang2023} or universal non-geodesic manifold \gls{rcg} \cite{Li2023b,Li2023c,Bartoli2023}.
	The former solves an unconstrained version of original problem by quasi-Newton methods, then projects the solution back to the Stiefel manifold.
	It has no guarantee of optimality and may suffer from numerical instability.
	The latter generalizes the conjugate gradient methods to any Riemannian manifold, which updates the feasible point by progressing along the conjugate direction and projecting back to the manifold at each step.
	In this section, we first discuss the existing non-geodesic \gls{rcg} method and its drawbacks, then propose a novel group-wise geodesic \gls{rcg} method that operates directly on the Stiefel manifold for faster convergence.

@@

		This universal \gls{rcg} method is applicable to optimization problems over arbitrary manifolds \cite{Absil2009,Pan2022d}.

@@

		For maximization problem with smooth objective $f$ and block-unitary constraint \eqref{eq:bd_ris}, the steps for \gls{bd}-\gls{ris} group $g$ at iteration $r$ are summarized below:

@@

	\begin{subsection}{Geodesic \glsfmtshort{rcg}}
		Before introducing geodesic \gls{rcg}, we revisit some basic concepts in differential geometry and Lie algebra.

		% \cite{Abrudan2008,Abrudan2009}.

		A Lie group is simultaneously a continuous group and a differentiable manifold.
		Lie algebra refers to the tangent space of the Lie group at the identity element.
		The exponential map acts as a bridge between the Lie algebra and Lie group, which allows one to recapture the local group structure using linear algebra techniques.
		The set of unitary matrices $\mathbb{U}^{L \times L}$ forms a Lie group $U(L)$ under multiplication, and the corresponding Lie algebra $\mathfrak{u}(L) \triangleq \mathcal{T}_{\mathbf{I}}\mathbb{U}^{L \times L} = \{\mathbf{M} \in \mathbb{C}^{L \times L} \mid \mathbf{M}^\mathsf{H} + \mathbf{M} = \mathbf{0}\}$ consists of skew-Hermitian matrices.
		A geodesic emanating from the identity with velocity $\mathbf{D} \in \mathfrak{u}(L)$ can be described by \cite{Edelman1998}
		\begin{equation}
			\mathbf{G}_\mathbf{I}(\mu) = \exp(\mu \mathbf{D}),
			\label{eq:geodesic_identity}
		\end{equation}
		where $\exp(\mathbf{A}) = \sum_{k=0}^\infty (\mathbf{A}^k/k!)$ is the matrix exponential and $\mu$ is the step size (i.e., magnitude of the tangent vector).
		Note that the right translation is an isometry in $U(L)$.
		During the optimization of group $g$, the geodesic evaluated at the identity \eqref{eq:geodesic_identity} should be translated to $\mathbf{\Theta}_g^{(r)}$ for successive updates \cite{Abrudan2008}
		\begin{equation}
			\mathbf{G}_g^{(r)}(\mu) = \mathbf{G}_\mathbf{I}(\mu) \mathbf{\Theta}_g^{(r)} = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)},
			\label{eq:geodesic_translated}
		\end{equation}
		while the Riemannian gradient evaluated at $\mathbf{\Theta}_g^{(r)}$ \eqref{eq:gradient_riemannian} should be translated back to the identity for exploiting the Lie algebra \cite{Abrudan2008}
		\begin{equation}
			\tilde{\nabla}_{\mathrm{R},g}^{(r)} = \nabla_{\mathrm{R},g}^{(r)} \mathbf{\Theta}_g^{(r)\mathsf{H}} = \nabla_{\mathrm{E},g}^{(r)} \mathbf{\Theta}_g^{(r)\mathsf{H}} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathrm{E},g}^{(r)\mathsf{H}}}.
			% $\mathcal{O}(2 L^3)$
			\label{eq:gradient_translated}
		\end{equation}
		After gradient translation, the deviation parameter and conjugate direction can be determined similarly to \eqref{eq:parameter_cg} and \eqref{eq:direction_cg}
		\begin{equation}
			\tilde{\gamma}_g^{(r)} = \frac{\mathrm{tr}\bigl((\tilde{\nabla}_{\mathrm{R},g}^{(r)} - \tilde{\nabla}_{\mathrm{R},g}^{(r-1)}) {\tilde{\nabla}_{\mathrm{R},g}^{(r)\mathsf{H}}}\bigr)}{\mathrm{tr}\bigl(\tilde{\nabla}_{\mathrm{R},g}^{(r-1)} {\tilde{\nabla}_{\mathrm{R},g}^{(r-1)\mathsf{H}}}\bigr)}. % $\mathcal{O}(2 N_\mathrm{S}^3 + N_\mathrm{S}^2 + 2 N_\mathrm{S})$
			\label{eq:parameter_cg_geodesic}
		\end{equation}
		\begin{equation}
			{\mathbf{D}}_g^{(r)} = \tilde{\nabla}_{\mathrm{R},g}^{(r)} + \tilde{\gamma}_g^{(r)} {\mathbf{D}}_g^{(r-1)},
			\label{eq:direction_cg_geodesic}
		\end{equation}
		The solution can thus be updated along the geodesic in a multiplicative rotational manner
		\begin{equation}
			\mathbf{\Theta}_g^{(r+1)} = \mathbf{G}_g^{(r)}(\mu) = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)},
			\label{eq:update_geodesic}
		\end{equation}
		where an appropriate $\mu$ may be obtained by the Armijo rule.
		To double the step size, one can simply square the rotation matrix instead of recomputing the matrix exponential, since $\exp^2(\mu \mathbf{D}_g^{(r)}) = \exp(2 \mu \mathbf{D}_g^{(r)})$.

@@

\begin{section}{Channel Singular Values Redistribution}
	In this section, we first provide a toy example to illustrate the channel shaping advantage of \gls{bd}-\gls{ris} architecture.
	Next, we numerically characterize the channel singular value region based on the proposed Algorithm \ref{ag:rcg}.
	Finally, we derive some analytical singular value bounds in specific scenarios.
	\begin{subsection}{Toy Example}\label{sc:toy_example}
		We first illustrate the channel shaping capabilities of different \gls{ris} models by a toy example.
		Consider a $2 \times 2 \times 2$ setup where the direct link is blocked.
		The diagonal \gls{ris} is modeled by $\mathbf{\Theta}_\mathrm{D} = \mathrm{diag}(e^{\jmath \theta_1}, e^{\jmath \theta_2})$ while the fully-connected \gls{ris} has 4 independent angular parameters
		\begin{equation}
			\mathbf{\Theta}_\mathrm{U} = e^{\jmath \phi} \begin{bmatrix}
				e^{\jmath \alpha} \cos \psi  & e^{\jmath \beta} \sin \psi   \\
				-e^{-\jmath \beta} \sin \psi & e^{-\jmath \alpha} \cos \psi
			\end{bmatrix}.
			\label{eq:unitary_ris}
		\end{equation}
		It is worth noting that $\phi$ has no impact on the singular value because $\mathrm{sv}(e^{\jmath \phi} \mathbf{A}) = \mathrm{sv}(\mathbf{A})$.
		To simplify the analysis, we also enforce symmetry $\mathbf{\Theta}_\mathrm{U} = \mathbf{\Theta}_\mathrm{U}^\mathsf{T}$ by $\beta = \pi / 2$ such that both architectures have the same number of optimization variables.
		\begin{figure}
			\centering
			\includegraphics[width=\columnwidth]{assets/simulation/singular_trend.eps}
			\caption{$2 \times 2 \times 2$ (no direct) channel singular value shaping by diagonal and symmetric fully-connected \gls{ris}.}
			\label{fg:singular_trend}
		\end{figure}
		Fig.~\ref{fg:singular_trend} shows the channel singular values achieved by an exhaustive grid search over $(\theta_1, \theta_2)$ for diagonal \gls{ris} and $(\alpha, \psi)$ for symmetric fully-connected \gls{ris}.
		In this particular example, we observe that both singular values can be manipulated up to $9\%$ using diagonal \gls{ris} and $42\%$ using symmetric \gls{bd}-\gls{ris}, despite both architectures have the same number of scattering elements and optimization variables.\footnote{However, \gls{bd}-\gls{ris} does require more reconfigurable components that interconnect the scattering elements.}
		A larger performance gap is expected when the symmetric constraint on \eqref{eq:unitary_ris} can be relaxed.
		This example shows \gls{bd}-\gls{ris} can provide a wider dynamic range of channel singular values and motivates further studies on channel shaping.
	\end{subsection}

@@

\begin{section}{Power Gain and Achievable Rate Maximization}\label{sc:power_rate}
	\begin{subsection}{Channel Power Gain}\label{sc:power}
		The \gls{mimo} channel power gain maximization problem is formulated with respect to the \gls{bd}-\gls{ris} scattering matrix
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:power_passive}}{\label{ob:power_passive}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
		\end{maxi!}
		which generalizes the case of \gls{siso} \cite{Shen2020a}, \gls{miso} \cite{Santamaria2023,Fang2023}, single-stream \gls{mimo} \cite{Nerini2023,Nerini2023b}, and direct link-blocked \gls{mimo} with fully-connected \gls{ris} \eqref{eq:scattering_fc_optimal}.
		The key of solving \eqref{op:power_passive} is to balance the additive and multiplicative space alignments.
		Interestingly, in terms of maximizing the inner product $\langle \mathbf{H}_\mathrm{D}, \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rangle$, \eqref{op:power_passive} is reminiscent of the weighted orthogonal Procrustes problem \cite{Gower2004}
		\begin{mini!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} - \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:weighted_orthogonal_procrustes}}{}
			\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I},}{\label{cs:power_block_unitary}}{}
		\end{mini!}
		which relaxes the block-unitary constraint \eqref{cs:power_block_unitary} to unitary but still has no trivial solution.
		One lossy transformation exploits the Moore-Penrose inverse and moves $\mathbf{\Theta}$ to one side of the product \cite{Bell2003}, formulating two standard orthogonal Procrustes problems
		\begin{mini!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{B}^\dagger \mathbf{H}_\mathrm{D} - \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2 \text{ or } \lVert \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\dagger - \mathbf{H}_\mathrm{B} \mathbf{\Theta} \rVert _\mathrm{F}^2}{\label{op:standard_orthogonal_procrustes}}{}
			\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I},}{}{}
		\end{mini!}
		which have global optimal solutions
		\begin{equation}
			\mathbf{\Theta} = \mathbf{U} \mathbf{V}^\mathsf{H},
			\label{eq:orthogonal_procrustes_solution}
		\end{equation}
		where $\mathbf{U}$ and $\mathbf{V}$ are respectively the left and right compact singular matrices of $\mathbf{H}_\mathrm{B}^\dagger \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\mathsf{H}$ or $\mathbf{H}_\mathrm{B}^\mathsf{H} \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\dagger$ \cite{Golub2013}.
		We emphasize that \eqref{eq:scattering_fc_optimal} and \eqref{eq:orthogonal_procrustes_solution} are valid fully-connected \gls{ris} solutions to \eqref{op:power_passive} when the direct link is absent and present, but the latter is neither optimal nor a generalization of the former due to the lossy transformation.

@@

\begin{section}{Conclusion}
	This paper analyzes the channel shaping capability of \gls{ris} in terms of singular values redistribution.
	We consider a general \gls{bd} architecture that allows elements within the same group to interact, enabling more sophisticated manipulation than diagonal \gls{ris}.
	This translates to a wider dynamic range (with better tradeoff) of singular values and significant power and rate gains, especially in large-scale \gls{mimo} systems.
	We characterize the Pareto frontiers of channel singular values via optimization approach and provide analytical bounds in rank-deficient and unitary scenarios.
	Specifically, the former is done by proposing an efficient \gls{rcg} algorithm for smooth \gls{bd}-\gls{ris} optimization problems, which offers better objective value and faster convergence than existing methods.
	We also present two beamforming designs for rate maximization problem, one based on alternating optimization for optimal performance and the other decouples the \gls{ris}-transceiver design for lower complexity.
	Extensive simulations show that the advantage of \gls{bd}-\gls{ris} stems from its superior space alignment and channel rearrangement capability, which scales with the number of elements, group size, \gls{mimo} dimensions, and channel diversity.

>>>>

	Second, we propose an efficient \gls{bd}-\gls{ris} design framework that solves general optimization problems (e.g., singular value redistribution and rate maximization) by geodesic\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} \gls{rcg}.
	% to numerically analyze the channel shaping capability of \gls{bd}-\gls{ris}, we exploit the Riemannian geometry of the Stiefel manifold and
	This method modified from \cite{Abrudan2008,Abrudan2009} exploits the Riemannian geometry of the Stiefel manifold and performs group-wise multiplicative rotational updates along the geodesics, which avoids retraction and facilitates step size selection.
	It thus converges faster than existing non-geodesic approaches \cite{Absil2009,Pan2022d}.
	This is the first work to tailor an efficient optimization framework for \gls{bd}-\gls{ris}.

@@

\begin{section}{\glsfmtshort{bd}-\glsfmtshort{ris} Design by Geodesic \glsfmtshort{rcg}}
	% In this section, we first provide an overview on signal processing techniques for group-connected \gls{bd}-\gls{ris} design problems, then propose a novel group-wise geodesic \gls{rcg} method that exploits the properties of unitary group and operates directly on the Stiefel manifold.
	% The proposed method is applicable to any smooth optimization problem on the Stiefel manifold and is particularly suitable for \gls{bd}-\gls{ris} design problems.
	% We will later show that the proposed method provides better objective value and faster convergence for smooth optimization problems of arbitrary group size.

	Many group-connected \gls{bd}-\gls{ris} design problems can be formulated in the following general form
	\begin{maxi!}
		{\scriptstyle{\mathbf{\Theta}}}{f(\mathbf{\Theta})}{\label{op:general}}{\label{ob:general}}
		\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I},}{\quad \forall g,}{\label{co:general_unitary}}
	\end{maxi!}
	where the objective function $f(\mathbf{\Theta})$ can be any function of the \gls{bd}-\gls{ris} scattering matrix $\mathbf{\Theta}$, for example, channel singular value (to be discussed in Section \ref{sc:singular_value}) and achievable rate (to be discussed in Section \ref{sc:power_rate}).
	The feasible domain of each group is a $L$-dimensional Stiefel manifold $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$, which is non-convex and non-Euclidean.
	Therefore, relevant optimization problems are usually solved by relax-then-project methods \cite{Fang2023} or universal non-geodesic manifold \gls{rcg} \cite{Li2023b,Li2023c,Bartoli2023}.
	The former solves an unconstrained problem \eqref{ob:general} by quasi-Newton methods, then projects the solution back to the Stiefel manifold.
	It has no guarantee of optimality and may suffer from numerical instability.
	The latter generalizes the conjugate gradient methods to any Riemannian manifold, which updates the feasible point by progressing along the conjugate direction and projecting back to the manifold at each step.
	In the following context, we first discuss the existing non-geodesic \gls{rcg} method and its drawbacks, then propose a novel group-wise geodesic \gls{rcg} method that operates directly on the Stiefel manifold for faster convergence.

@@

		A geodesic is a curve representing the shortest path between two points in a Riemannian manifold, whose tangent vectors remain parallel when transporting along the curve.
		The universal \gls{rcg} method proposed in \cite{Absil2009,Pan2022d} is non-geodesic and applicable to optimization problems over arbitrary manifolds.

@@

		For optimization problem \eqref{op:general}, the steps for \gls{bd}-\gls{ris} group $g$ at iteration $r$ are summarized below:

@@

	\begin{subsection}{Geodesic \glsfmtshort{rcg}}
		Before introducing geodesic \gls{rcg}, we revisit some basic concepts in differential geometry and Lie algebra.
		% \cite{Abrudan2008,Abrudan2009}.
		A Lie group is simultaneously a continuous group and a differentiable manifold.
		Lie algebra refers to the tangent space of the Lie group at the identity element.
		The exponential map acts as a bridge between the Lie algebra and Lie group, which allows one to recapture the local group structure using linear algebra techniques.
		The set of unitary matrices $\mathbb{U}^{L \times L}$ forms a Lie group $U(L)$ under multiplication, and the corresponding Lie algebra $\mathfrak{u}(L) \triangleq \mathcal{T}_{\mathbf{I}}\mathbb{U}^{L \times L} = \{\mathbf{M} \in \mathbb{C}^{L \times L} \mid \mathbf{M}^\mathsf{H} + \mathbf{M} = \mathbf{0}\}$ consists of skew-Hermitian matrices.
		A geodesic emanating from the identity with velocity $\mathbf{D} \in \mathfrak{u}(L)$ can be described by \cite{Edelman1998}
		\begin{equation}
			\mathbf{G}_\mathbf{I}(\mu) = \exp(\mu \mathbf{D}),
			\label{eq:geodesic_identity}
		\end{equation}
		where $\exp(\mathbf{A}) = \sum_{k=0}^\infty (\mathbf{A}^k/k!)$ is the matrix exponential and $\mu$ is the step size (i.e., magnitude of the tangent vector).
		Note that the right translation is an isometry in $U(L)$.
		During the optimization of group $g$, the geodesic evaluated at the identity \eqref{eq:geodesic_identity} should be translated to $\mathbf{\Theta}_g^{(r)}$ for successive updates \cite{Abrudan2008}
		\begin{equation}
			\mathbf{G}_g^{(r)}(\mu) = \mathbf{G}_\mathbf{I}(\mu) \mathbf{\Theta}_g^{(r)} = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)},
			\label{eq:geodesic_translated}
		\end{equation}
		while the Riemannian gradient evaluated at $\mathbf{\Theta}_g^{(r)}$ \eqref{eq:gradient_riemannian} should be translated back to the identity for exploiting the Lie algebra \cite{Abrudan2008}
		\begin{equation}
			\tilde{\nabla}_{\mathrm{R},g}^{(r)} = \nabla_{\mathrm{R},g}^{(r)} \mathbf{\Theta}_g^{(r)\mathsf{H}} = \nabla_{\mathrm{E},g}^{(r)} \mathbf{\Theta}_g^{(r)\mathsf{H}} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathrm{E},g}^{(r)\mathsf{H}}}.
			% $\mathcal{O}(2 L^3)$
			\label{eq:gradient_translated}
		\end{equation}
		After gradient translation, the deviation parameter and conjugate direction can be determined similarly to \eqref{eq:parameter_cg} and \eqref{eq:direction_cg}
		\begin{equation}
			\tilde{\gamma}_g^{(r)} = \frac{\mathrm{tr}\bigl((\tilde{\nabla}_{\mathrm{R},g}^{(r)} - \tilde{\nabla}_{\mathrm{R},g}^{(r-1)}) {\tilde{\nabla}_{\mathrm{R},g}^{(r)\mathsf{H}}}\bigr)}{\mathrm{tr}\bigl(\tilde{\nabla}_{\mathrm{R},g}^{(r-1)} {\tilde{\nabla}_{\mathrm{R},g}^{(r-1)\mathsf{H}}}\bigr)}. % $\mathcal{O}(2 N_\mathrm{S}^3 + N_\mathrm{S}^2 + 2 N_\mathrm{S})$
			\label{eq:parameter_cg_geodesic}
		\end{equation}
		\begin{equation}
			{\mathbf{D}}_g^{(r)} = \tilde{\nabla}_{\mathrm{R},g}^{(r)} + \tilde{\gamma}_g^{(r)} {\mathbf{D}}_g^{(r-1)},
			\label{eq:direction_cg_geodesic}
		\end{equation}
		The solution can thus be updated along the geodesic in a multiplicative rotational manner
		\begin{equation}
			\mathbf{\Theta}_g^{(r+1)} = \mathbf{G}_g^{(r)}(\mu) = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)},
			\label{eq:update_geodesic}
		\end{equation}
		where an appropriate $\mu$ may be obtained by the Armijo rule.
		To double the step size, one can simply square the rotation matrix instead of recomputing the matrix exponential, since $\exp^2(\mu \mathbf{D}_g^{(r)}) = \exp(2 \mu \mathbf{D}_g^{(r)})$.

@@

\begin{section}{Channel Singular Values Redistribution}
	\label{sc:singular_value}
	In this section, we first provide a toy example to illustrate the channel shaping advantage of \gls{bd}-\gls{ris} architecture.
	Next, we numerically characterize the channel singular value region based on the proposed Algorithm \ref{ag:rcg}.
	Finally, we derive some analytical singular value bounds in specific scenarios.
	\begin{subsection}{Toy Example}\label{sc:toy_example}
		We first illustrate the channel shaping capabilities of different \gls{ris} models by a toy example.
		Consider a $2 \times 2 \times 2$ setup where the direct link is blocked.
		The diagonal \gls{ris} is modeled by $\mathbf{\Theta}_\mathrm{D} = \mathrm{diag}(e^{\jmath \theta_1}, e^{\jmath \theta_2})$ while the fully-connected \gls{ris} has 4 independent angular parameters
		\begin{equation}
			\mathbf{\Theta}_\mathrm{U} = e^{\jmath \phi} \begin{bmatrix}
				e^{\jmath \alpha} \cos \psi  & e^{\jmath \beta} \sin \psi   \\
				-e^{-\jmath \beta} \sin \psi & e^{-\jmath \alpha} \cos \psi
			\end{bmatrix}.
			\label{eq:unitary_ris}
		\end{equation}
		It is worth noting that $\phi$ has no impact on the singular value because $\mathrm{sv}(e^{\jmath \phi} \mathbf{A}) = \mathrm{sv}(\mathbf{A})$.
		To simplify the analysis, we also enforce symmetry $\mathbf{\Theta}_\mathrm{U} = \mathbf{\Theta}_\mathrm{U}^\mathsf{T}$ by $\beta = \pi / 2$ such that both architectures have the same number of optimization variables.
		\begin{figure}
			\centering
			\includegraphics[width=\columnwidth]{assets/simulation/singular_trend.eps}
			\caption{$2 \times 2 \times 2$ (no direct) channel singular value shaping by diagonal and symmetric fully-connected \gls{ris}.}
			\label{fg:singular_trend}
		\end{figure}
		Fig.~\ref{fg:singular_trend} shows the channel singular values achieved by an exhaustive grid search over $(\theta_1, \theta_2)$ for diagonal \gls{ris} and $(\alpha, \psi)$ for symmetric fully-connected \gls{ris}.
		In this particular example, we observe that both singular values can be manipulated up to $9\%$ using diagonal \gls{ris} and $42\%$ using symmetric \gls{bd}-\gls{ris}, despite both architectures have the same number of scattering elements and optimization variables.\footnote{However, \gls{bd}-\gls{ris} does require more reconfigurable components that interconnect the scattering elements.}
		A larger performance gap is expected when the symmetric constraint on \eqref{eq:unitary_ris} can be relaxed.
		This example shows \gls{bd}-\gls{ris} can provide a wider dynamic range of channel singular values and motivates further studies on channel shaping.
	\end{subsection}

@@

\begin{section}{Power Gain and Achievable Rate Maximization}
	\label{sc:power_rate}
	\begin{subsection}{Channel Power Gain}\label{sc:power}
		The \gls{mimo} channel power gain maximization problem is formulated with respect to the \gls{bd}-\gls{ris} scattering matrix
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:power_passive}}{\label{ob:power_passive}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
		\end{maxi!}
		which generalizes the case of \gls{siso} \cite{Shen2020a}, \gls{miso} \cite{Santamaria2023,Fang2023}, single-stream \gls{mimo} \cite{Nerini2023,Nerini2023b}, and direct link-blocked \gls{mimo} with fully-connected \gls{ris} \eqref{eq:scattering_fc_optimal}.
		The key of solving \eqref{op:power_passive} is to balance the additive and multiplicative space alignments.
		Interestingly, in terms of maximizing the inner product $\langle \mathbf{H}_\mathrm{D}, \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rangle$, \eqref{op:power_passive} is reminiscent of the weighted orthogonal Procrustes problem \cite{Gower2004}
		\begin{mini!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} - \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:weighted_orthogonal_procrustes}}{}
			\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I},}{\label{cs:power_block_unitary}}{}
		\end{mini!}
		which relaxes the block-unitary constraint \eqref{cs:power_block_unitary} to unitary but still has no trivial solution.
		One lossy transformation exploits the Moore-Penrose inverse and moves $\mathbf{\Theta}$ to one side of the product \cite{Bell2003}, formulating two standard orthogonal Procrustes problems
		\begin{mini!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{B}^\dagger \mathbf{H}_\mathrm{D} - \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2 \text{ or } \lVert \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\dagger - \mathbf{H}_\mathrm{B} \mathbf{\Theta} \rVert _\mathrm{F}^2}{\label{op:standard_orthogonal_procrustes}}{}
			\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I},}{}{}
		\end{mini!}
		which have global optimal solutions
		\begin{equation}
			\mathbf{\Theta} = \mathbf{U} \mathbf{V}^\mathsf{H},
			\label{eq:orthogonal_procrustes_solution}
		\end{equation}
		where $\mathbf{U}$ and $\mathbf{V}$ are respectively the left and right compact singular matrices of $\mathbf{H}_\mathrm{B}^\dagger \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\mathsf{H}$ or $\mathbf{H}_\mathrm{B}^\mathsf{H} \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\dagger$ \cite{Golub2013}.
		We emphasize that \eqref{eq:scattering_fc_optimal} and \eqref{eq:orthogonal_procrustes_solution} are valid fully-connected \gls{ris} solutions to \eqref{op:power_passive} when the direct link is absent and present, but the latter is neither optimal nor a generalization of the former due to the lossy transformation.

@@

\begin{section}{Conclusion}
	This paper analyzes the channel shaping capability of \gls{ris} in terms of singular values redistribution.
	We consider a general \gls{bd} architecture that allows elements within the same group to interact, enabling more sophisticated manipulation than diagonal \gls{ris}.
	This translates to a wider dynamic range (with better tradeoff) of singular values and significant power and rate gains, especially in large-scale \gls{mimo} systems.
	We characterize the Pareto frontiers of channel singular values via optimization approach and provide analytical bounds in rank-deficient and unitary scenarios.
	Specifically, the former is done by proposing an efficient \gls{rcg} algorithm for \gls{bd}-\gls{ris} optimization problems, which offers better objective value and faster convergence than existing methods.
	We also present two beamforming designs for rate maximization problem, one based on alternating optimization for optimal performance and the other decouples the \gls{ris}-transceiver design for lower complexity.
	Extensive simulations show that the advantage of \gls{bd}-\gls{ris} stems from its superior space alignment and channel rearrangement capability, which scales with the number of elements, group size, \gls{mimo} dimensions, and channel diversity.
